---
title: "mbta_615final"
author: "Danya Zhang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(echo = FALSE, tidy.opts=list(width.cutoff=70), tidy=TRUE,
                      warning = FALSE)
library(dplyr)
```

\normalsize The MBTA, more commonly known as the "T" by locals, is the first 
\normalsize subway system ever implemented in the United States. It was voted 
\normalsize into law in 1964, serving the greater Boston area. Some time later, 
\normalsize a consulting firm was hired to consolidate the various transit lines; 
\normalsize it was then that the MBTA system became as we know it today with the 
\normalsize color-coded lines. The data for this report can be downloaded
\normalsize here: 
\normalsize https://mbta-massdot.opendata.arcgis.com/datasets/mbta-travel-times-2022/about
\normalsize For this report, I've selected the Q1 LR dataset.
$\\$
$\\$
\normalsize In this report, we would like to examine the the reliability of the MBTA
\normalsize system using statistics and visualizations to see if departure and 
\normalsize arrival times are accurate. Let's first read in the data. To make it 
\normalsize easier, since the data is quite large, we will work with a subset 
\normalsize of the data which represents a week of data.
$\\$
```{r}
#subsetting data to the first week of 2022
data <- read.csv("/Users/dz/Documents/MSSP/GitHub/MA615 final/Final-Project-MBTA/TravelTimes_2022/2022-Q1_LRTravelTimes.csv")
df <- as.data.frame(table(data$service_date))
sum <- sum(df[c(1:7),2]) 
data <- head(data, sum)
```

\Large Getting to know the data
$\\$
$\\$
\large Summary Statistics
$\\$
$\\$
\normalsize Let's look at some summary statistics of the data.
```{r}
#summary statistics 
stats <- data %>%
  group_by(from_stop_id, to_stop_id) %>%
  summarise(max = max(travel_time_sec), min = min(travel_time_sec),
            mean = mean(travel_time_sec), sd = sd(travel_time_sec))
head(stats)
```

\normalsize There were some travel times that are clearly impossible, perhaps 
\normalsize due to a data entry error. So we will delete any observations with 
\normalsize less than a 10 second travel time. For the visualizations, we will 
\normalsize further subset the data to a single from_stop_id, to_stop_id, 
\normalsize direction_id, route_id, and direction_id (subset1). 
$\\$
$\\$
\large Data Cleaning
```{r, echo=TRUE}
#some travel times that are clearly impossible. possible data entry error
short_ind <- data[which(data$travel_time_sec<10),]
remove_ind <- as.numeric(dimnames(short_ind)[[1]])
data <- data[-remove_ind,]

#subsetting based on the same from_stop_id, to_stop_id, route_id, direction_id
subset1 <- data %>%
  filter(from_stop_id == 70134 &
           to_stop_id == 170136 &
           route_id == "Green-B" &
           direction_id == 1)

#first week of 2022
subset1 <- as.data.frame(subset1)
```

\normalsize Now, let's visualize the data using some graphs. $\\$
$\\$
\Large Visualizations
$\\$
$\\$
\large Histogram
```{r, echo=FALSE}
#histogram
hist(subset1$travel_time_sec, xlab = "Travel Time (sec)", 
main = "Frequency of MBTA travel times")
```
$\\$
\normalsize Aside from the very few outliers, the histogram is otherwise fairly 
\normalsize bell-shaped. This suggest that travel times are roughly Gaussian 
\normalsize and the MBTA travel times between the two stops are fairly quick. 
$\\$

\large Boxplot
```{r, echo=FALSE}
library(ggplot2)
#boxplot
subset1 %>%
ggplot(aes(x = as.Date(service_date), y = travel_time_sec/60, group=service_date)) +
  geom_boxplot() +
  labs(title = "Boxplot of travel times", x = "Date (2022)",y = "Travel Time (min)",
       subtitle = "stop 70134 to stop 170136, greenline-B, direction_id 1") +
  scale_x_date(date_breaks = "1 day", date_label="%b%d") +
  theme_bw()

#looks pretty consistent. There are outliers but this could be amounted to 
#vehicle malfunctions, accidents, emergencies, unruly passengers, etc.
```
\normalsize As the box plots are roughly on the same horizontal level as each 
\normalsize other, there seems to be little to no variation. Ignoring the outliers, 
\normalsize the medians, 25% quantile and 75% quantile are at roughly the same 
\normalsize level. This is good as it means the Greenline-B is very consistent 
\normalsize with travel times.

$\\$
\large Scatter plot
$\\$
\normalsize Now, let's use a scatter plot to explore possible variation between 
\normalsize two extremes. Here I have selected two very different days in terms 
\normalsize of passenger traffic: Sunday and Tuesday. Let's see if there is a 
\normalsize difference between weekdays and weekends.
$\\$
```{r, echo=FALSE}
subset1_tues <- subset1[which(subset1$service_date=="2022-01-04"),]
subset1_sun <- subset1[which(subset1$service_date=="2022-01-02"),]
subset1_suntues <- rbind(subset1_tues, subset1_sun)

ggplot(subset1_suntues, aes(start_time_sec/60/60, end_time_sec/60/60, group=service_date,
                           color=service_date)) +
  geom_point() +
  labs(title = "End Time vs. Start Time", x="Start time (hr)", 
       y="End time (hr)",subtitle = "stop 70134 to stop 170136, greenline-B, direction_id 1") +
  guides(color = guide_legend(title = "Service Date")) +
  theme_bw()
```
\normalsize Scatter plots appear linear and stacked on top of each other, which 
\normalsize means start and end times are consistent between days, meaning the  
\normalsize MBTA is fairly reliable.
$\\$
$\\$

\Large Hypotheses Tests 
$\\$
$\\$
\normalsize Moving on to the hypotheses tests, we'd like to use some statistical 
\normalsize tests to see if MBTA travel times are reliable and consistent.
$\\$
$\\$
\large Anova test 
$\\$
\normalsize Let's start with an ANOVA test. Our null and alternative hypotheses 
\normalsize are as follows: 
$\\$ $\\$
$H_0 =$ there are no significant differences in travel times $\\$
$H_1$ = there are significant differences in travel times
```{r}

library(rstatix)
anova <- data %>% anova_test(travel_time_sec ~ service_date)
anova
```
\normalsize The p-value for the anova test is very small this shows that 
\normalsize we reject our null hypotheses, meaning that there are significant
\normalsize differences in travel times. This might be surprising considering
\normalsize the graphs above did not show this. However, it is very important to know
\normalsize that p-values do not represent a definitive answer. Let's consider 
\normalsize another type of test: the paired t-test. $\\$

\large Paired T-test
```{r}
ttest <- t_test(data=subset1_suntues,travel_time_sec ~ service_date)
ttest2 <- t_test(data=subset1_suntues, start_time_sec ~ service_date)
ttest3 <- t_test(data=subset1_suntues, end_time_sec ~ service_date)
ttest
ttest2
ttest3
```

\normalsize The paired t-test above compares travel times, start times, and end,
\normalsize times between Sunday and Tuesday. We'd like to discern whether these
\normalsize variables differ based on days. The alternate and null 
\normalsize hypotheses remain the same as above. Looking at the results, it seems \normalsize that only the first test for travel time rejects the null 
\normalsize but only at a 0.1 significance level; the other two tests for start
\normalsize time and end time fail to reject the null at a 0.01, 0.05, and 0.1 
\normalsize level, therefore we can make a general conclusion that there is \normalsize little to no difference between days and the MBTA is fairly \normalsize consistent with minor variation in travel times. 
$\\$

\normalsize This concludes our report on the reliability of the MBTA greenline.
\normalsize It appears that the overall, the greenline is fairly reliable.
\normalsize From personal experience, as someone who has gone to school in \normalsize Boston for four years, the MBTA's senescence is obvious. It's
\normalsize quite slow and loud, at least the subway is. A part of me would
\normalsize like to rebute the fact that the greenline is reliable, however
\normalsize the predictive LCD screens around the city are accurate to say the 
\normalsize least. Overall, the MBTA greenline is fairly reliable. 
$\\$
\normalsize github repo: https://github.com/GRS-MA615/Final-Project-MBTA

$\\$
```{r}
#some demo code for r shiny application
# library(mapsapi)
# key <- "AIzaSyBb1P6Czwie4KeT0BAKlHLoCzzUXRh9qyA"
# 
# #580 to prudential
# doc <- mp_directions(origin = "Boston University",
#                      destination = "Newbury Street",
#                      alternatives = TRUE,
#                      key = key,
#                      quiet = TRUE,
#                      transit_mode = c("bus", "subway"))
# 
# r <- mp_get_routes(doc)
# 
# #map visualization
# library(leaflet)
# pal <- colorFactor(palette = "Dark2", domain = r$alternative_id)
# leaflet() %>%
#   addProviderTiles("CartoDB.DarkMatter") %>%
#   addPolylines(data = r, opacity = 1, weight = 7, color = ~pal(alternative_id))
```



